Appendix: Why the complement of AUC isn't a casewise CV criterion

### 2024-07-20 scratch/notes_georges/JSS-article/JSS-article-4_corrected_typos_GM.Rmd
JSS-article-4 with modifications

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0003-0076-5532
    affiliation: 'York University'
    address: |
      | Toronto, Ontario, Canada
    email: \email{georges@yorku.ca}
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validating Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validating Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validating Regression Models"
abstract: >
  We describe the \pkg{cv} package, which implements cross-validation for standard \proglang{R} regression models through a uniform and simple to use generic function. Methods for the \code{cv()} function are provided for many commonly employed classes of statistical models, including mixed-effects models, and it is straightforward to extend the \pkg{cv} package by writing methods for additional classes of models. The \code{cv()} function also can cross-validate complex model-selection procedures, such as those that include variable transformations, predictor selection, and selection among competing models. The \code{cv()} function generally supports parallel computations, and the supplied methods for linear and generalized linear models take advantage of computational efficiencies.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv.bib"]
---

@@

In the interest of brevity, we won't describe all of the features of the \pkg{cv} package here, concentrating on the aspects of the package that are relatively novel. For example, the \code{cv()} function can perform computations in parallel and can independently replicate a cross-validation procedure several times. There are also data management facilities in the package, such as coercing the objects produced by the \code{cv()} function into data frames for further analysis. These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provides greater detail on some of topics that we do describe, such as extensions to the package.

@@

The most straightforward way to implement cross-validation in \proglang{R} for statistical modeling functions that are written in the canonical manner is to use \code{update()} to refit the model with each fold removed. This is the approach taken in the default method for \code{cv()}, and it is appropriate if the cases are independently sampled. Refitting the model in this manner for each fold is generally feasible when the number of folds in modest, but can be prohibitively costly for leave-one-out cross-validation when the number of cases is large.

@@

[^WLS]: This is a definitional formula, which assumes that the model matrix $\mathbf{X}$ is of full column rank, and which can be subject to numerical instability when $\mathbf{X}$ is ill-conditioned. \code{lm()} uses the singular-value decomposition of the model matrix to obtain computationally more stable results.

@@

To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identify [@Hager:1989],
$$
\left(\mathbf{A}_{m \times m} + \mathbf{U}_{m \times k} 
\mathbf{C}_{k \times k} \mathbf{V}_{k \times m} \right)^{-1} = \mathbf{A}^{-1} - \mathbf{A}^{-1}\mathbf{U} \left(\mathbf{C}^{-1} + 
\mathbf{VA}^{-1}\mathbf{U} \right)^{-1} \mathbf{VA}^{-1}
$$
where $\mathbf{A}$ is a nonsingular order-$n$ matrix. We apply this result by letting
\begin{align*}
	\mathbf{A} &= \mathbf{X}^\top \mathbf{W} \mathbf{X} \\
	\mathbf{U} &= \mathbf{X}_\mathbf{j}^\top \\
	\mathbf{V} &= - \mathbf{X}_\mathbf{j} \\
	\mathbf{C} &= \mathbf{W}_\mathbf{j} \\
\end{align*}
where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^\top$ represents the vector of indices for the cases in the $j$th fold, $j = 1, \ldots, k$. The negative sign in $\mathbf{V} = - \mathbf{X}_\mathbf{j}$ reflects the *removal*, rather than addition, of the cases in $\mathbf{j}$.

>>>>

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0003-0076-5532
    affiliation: 'York University'
    address: |
      | Toronto, Ontario, Canada
    email: \email
    url: http://blackwell.math.yorku.ca/gmonette
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validating Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validating Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validating Regression Models"
abstract: >
  We describe the \pkg{cv} package, which implements cross-validation for standard \proglang{R} regression models through a uniform and simple to use generic function. Methods for the \code{cv()} function are provided for many commonly employed classes of statistical models, including mixed-effects models, and it is straightforward to extend the \pkg{cv} package by writing methods for additional classes of models. The \code{cv()} function also can cross-validate complex model-selection procedures, such as those that include variable transformations, predictor selection, and selection among competing models. The \code{cv()} function generally supports parallel computations, and the supplied methods for linear and generalized linear models take advantage of computational efficiencies.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv.bib"]
---

@@

In the interest of brevity, we won't describe all of the features of the \pkg{cv} package here, concentrating on the aspects of the package that are relatively novel. For example, the \code{cv()} function can perform computations in parallel and can independently replicate a cross-validation procedure several times. There are also data management facilities in the package, such as coercing the objects produced by the \code{cv()} function into data frames for further analysis. These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provide greater detail on some topics that we do describe, such as extensions to the package.

@@

The most straightforward way to implement cross-validation in \proglang{R} for statistical modeling functions that are written in the canonical manner is to use \code{update()} to refit the model with each fold removed. This is the approach taken in the default method for \code{cv()}, and it is appropriate if the cases are independently sampled. Refitting the model in this manner for each fold is generally feasible when the number of folds is modest, but can be prohibitively costly for leave-one-out cross-validation when the number of cases is large.

@@

[^WLS]: This is a definitional formula, which assumes that the model matrix $\mathbf{X}$ is of full column rank, and which can be subject to numerical instability when $\mathbf{X}$ is ill-conditioned. \code{lm()} uses the QR decomposition of the model matrix with pivoting to obtain computationally more stable results.

@@

To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identity [@Hager:1989],
$$
\left(\mathbf{A}_{m \times m} + \mathbf{U}_{m \times k} 
\mathbf{C}_{k \times k} \mathbf{V}_{k \times m} \right)^{-1} = \mathbf{A}^{-1} - \mathbf{A}^{-1}\mathbf{U} \left(\mathbf{C}^{-1} + 
\mathbf{VA}^{-1}\mathbf{U} \right)^{-1} \mathbf{VA}^{-1}
$$
where $\mathbf{A}$ is a nonsingular order-$n$ matrix. We apply this result by letting
\begin{align*}
	\mathbf{A} &= \mathbf{X}^\top \mathbf{W} \mathbf{X} \\
	\mathbf{U} &= \mathbf{X}_\mathbf{j}^\top \\
	\mathbf{V} &= - \mathbf{X}_\mathbf{j} \\
	\mathbf{C} &= \mathbf{W}_\mathbf{j} \\
\end{align*}
where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^\top$ represents the vector of indices for the cases in the $j$th fold, $j = 1, \ldots, k$. The negative sign in $\mathbf{V} = - \mathbf{X}_\mathbf{j}$ reflects the *removal*, rather than addition, of the cases in $\mathbf{j}$.
